One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-10-14 doc/response.tex

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\breview}{\begin{quotation}\begin{bf}\noindent}
\newcommand{\ereview}{\end{bf}\end{quotation}}
\newcommand{\reviewbullet}[1]{\breview \begin{itemize}\item #1 \end{itemize}\ereview}
\begin{document}

Dear editors:

We have attached a revised version of our manuscript ``A simple null model for inferences from network enrichment analysis'', and a diff
against our original manuscript. We have also answered the reviewer's comments
below.

In general, by reading the reviewers comments, we feel that we failed to deliver the purpose of the manuscript to the reviewer, who indeed appears to have misunderstood what we are aiming at with our manuscript. We have hence added clarifications to the introduction and the discussion. We hope that this will make the story even more interesting.

@@

Yours Sincerely,\\[1.5cm]
Gustavo Jeuken and
Lukas K\"{a}ll
\\

\section*{Reviewer \#1}
\breview
Jeunken and Käll describe a method for generating a null distribution for so called ``network enrichment analyses'', which are sometimes framed as ``crosstalk'' analyses. The only data presented is showing that the method generates accurate pvalues under the null.
\medskip
As far as I can tell the work is technically correct. Their statement of the null hypothesis they are testing is sufficiently precise. I agree with the authors that their conceptualization of the problem is simpler than methods like BinoX. But I disagree that the interpretation can be assumed to be meaningful, and the authors provide no evidence.
\ereview
The reviewer points out that the manuscript only discusses the calibration of $p$ values, and never actually demonstrates the utility of such $p$ values. The reviewer even does not find the findings of the manuscript meaningful. While it is worth noting that  according to the Criteria For Publication PLOS ONE, (\url{https://journals.plos.org/plosone/s/criteria-for-publication}) perceived meaningfulness is not a publication criterion for a manuscript, lack of meaning is obviously an undesirable property of any scientific report. We have carefully rewritten the introduction to make the readers get a sense of the underlying conflict driving our manuscript. We feel ourselves limited from writing that existing methods for network enrichment analysis (NEA) produce grossly inflated statistics. We instead have tried to, in a constructive manner, described a method for deriving statistics that is natural from a users perspective. This model could be integrated into any of the currently used  models for NEA. It is an open question if one could demonstrate improved sensitivity with NEA over over-representation analysis in such implementations.

\breview
They claim that BinoX yields very inaccurate p-values. If the results yielded by the proposed method are basically the same as BinoX (in terms of ranks), then there is no major problem. But the authors state that they don’t want to get into evaluating ``performance in terms of sensitivity''. I find this disingenuous; they have to at least show how their ranked results compare to BinoX on real data and do some kind of benchmarking. This is important because I believe there is a difference in the null hypothesis stated in the current work and that used by BinoX that makes the comparison presented here unfair -- they are not testing the same thing.
\ereview

We fully agree with the reviewer that the null of BinoX is different that GeneSetDP. In fact that is the purpose of our manuscript. The reason why no sensitivity test is made is of course that there is no way to compete in sensitivity against a method with inflated statistics. If the yardstick is flawed no comparison can be made. Here we establish a fair yardstick to compare against. In our manuscript we are not trying to sell a new method for NEA, we are merely trying to convince the field that they should report statistics that makes sense to a user.

\breview
The crux is that the method proposed here focuses on the properties of the query genes; randomly sampling query genes makes the proposed method a test of whether the query genes are unusual, not necessarily specifically with respect to a particular pathway. For example, if the query contains a hub gene (assuming hubs are rare) then it might be found ``significantly'' connected to many pathways. The ``feature unlikely to occur by chance'' is that the query contains a hub gene, not that it connects to many pathways. To state it in an extreme way, in this situation biologist should skip the pathway analysis and just take note of the fact that the query includes a hub. Using the authors’ method without this insight could lead to very misleading conclusions.

@@

\medskip

As I understand it, BinoX tries to account for these difficulties by modeling realizations of the network that retain key topological properties like node degree distribution while holding the query constant (in effect). The current authors criticize the BinoX approach, but ignore the problem entirely. A closer (but not very close) approximation of the BinoX approach might be, for the Monte Carlo method proposed here, to bias the sampling of Q’ from G such that the distribution of the node degrees of the genes in Q’ resembles the distribution of the actual query set of interest Q. In any case, I believe that the BinoX approach is not directly comparable. The relationship needs to be discussed and some demonstration of the method yielding useful and specific results is needed. If the authors cannot do this then they must remove the discussion of BinoX. Without that, and without any benchmarking, the paper is essentially about an algorithm to compute N(s), with no biological content or relevance.

\ereview

Here, we respectfully disagree with the reviewer. The reviewer's perspective do not reflect the actual use case. The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in respect to a pathway database.  In this question, the introduction of a  network is not relevant more than in that it might shift the score distribution against pathways. A network may or may not help to increase the sensitivity of ORA methods, but it is better left out of the null hypothesis.  A user should not have to worry about the connectivity of individual genes. Just as the reviewer points out, there are certain hub genes that are connected to more pathways than others through any network. This will be reflected by the score distribution produced by our described methods. If the query set contains such a gene, the query is more likely to be significant against a pathway, however that is the desired property of the methods. Conversely, a query containing one or more genes not connected to any pathways is less likely to be significantly enriched. This is the nature of this type of analysis.

However, a perhaps more problematic fact is that the methods in this manuscript, just in all NEA methods, test every pathway independently for significance, and do not evaluate correlation between different pathways.
\breview

\breview
Minor: There are some typos I noted such as 'litterature' and 'posibile'.
\ereview
We updated these typos, and we also overall improved the language of the manuscript.

### 2018-10-15 doc/genesetdp.tex
More spell check
Simplified preamble
grammar check

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

@@

A limiting factor of ORA is the quality of the databases defining pathways. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. While the terminology varies in the literature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).

@@

Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need of parametric score distributions for the calculatios of significance.
Combining the GeneSetDP and  GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

@@

We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$: ``There are not more links between the query and pathway gene sets than expected by chance''. Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.

In effect, NEA is an attempt to increase the sensitivity of ORA. The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over-representation in respect to a pathway database. This goal is the same as for users of ORA. Hence the null model should be of the same nature for NEA and ORA. GeneSetDP and GeneSetMC are two algorithms that prove that one in NEA can test the uniqueness of the query set in respect to a pathway, just as done in ORA.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{changepage}
\usepackage[utf8x]{inputenc}
\usepackage{textcomp,marvosym}
\usepackage{cite,nameref,hyperref}
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

\usepackage[right]{lineno}
\usepackage{array}
\newcolumntype{+}{!{\vrule width 2pt}}

@@

A limiting factor of ORA is the quality of the databases defining pathways. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence, efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. While the terminology varies in the literature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).

@@

Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need of parametric score distributions for the calculations of significance.
Combining the GeneSetDP and  GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

@@

We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$: ``There are not more links between the query and pathway gene sets than expected by chance''. Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use for calculating accurate statistics. Generally, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.

In effect, NEA is an attempt to increase the sensitivity of ORA. The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over-representation in respect to a pathway database. This goal is the same as for users of ORA. Hence, the null model should be of the same nature for NEA and ORA. GeneSetDP and GeneSetMC are two algorithms that prove that one in NEA can test the uniqueness of the query set in respect to a pathway, just as done in ORA.
